% =====================================================================
%  TRB 2027 · MAIN PAPER SKELETON (draft)
%  连续动作空间下可证明 COLREGs【方向】合规的无人船避碰 · 投影式安全盾 + RL
%
%  这是把论文从模板占位文(0626-paper/paper.tex = Lorem ipsum)推进成【真骨架】的草稿。
%  每节：已定稿处写草稿正文；待实验数据处标 [PENDING: ...]；诚实边界处标 [HONESTY]。
%  可证明层正文由独立文件提供：\input{sec_provable_safety_layer}
%
%  诚实红线（CLAUDE.md §0 + 三份理论稿·全文必须守）：
%    绝不写 provably collision-free / provably compliant（裸）/ complete provable coverage。
%    头条能 claim = 连续动作 ∩ 可证明【方向】合规 ∩ COLREGs + 远场单步可证明 + 迫近不可避 sound 证书。
%    可证明档 = 中。到达率【不进卖点清单】（L192 B1 配平后无可证优势）。
%  TRB 硬约束：~7500 词·无附录（完整证明移交外部技术报告）。
% =====================================================================
\documentclass[12pt,letterpaper]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amsthm,amssymb,bm}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{hyperref}

\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{assumption}{Assumption}
\theoremstyle{remark}\newtheorem*{scope}{Scope and honest limits}

\title{Continuous-Action, Provably COLREGs-Direction-Compliant
Collision Avoidance for Unmanned Surface Vessels via a
Projection-Based Safety Shield}
\author{[authors]}
\date{}

\begin{document}
\maketitle

% =====================================================================
\section{Abstract}
% [DRAFT · 守诚实红线 · ~200 words target]
Provably COLREGs-\emph{direction}-compliant collision avoidance for unmanned
surface vessels (USVs) has, until now, forced a choice: either discretize the
action space and mask illegal actions (sacrificing continuous control), or keep
continuous actions but revert to soft reward shaping (sacrificing the guarantee).
We place a projection-based safety shield on maritime COLREGs directly in the
\emph{continuous} action space: a reinforcement-learning policy proposes a
continuous command, a rule state machine classifies the encounter regime and the
compliant give-way direction, and a quadratic program projects the command onto
the nearest action that is both direction-compliant and (single-step)
collision-free. We characterize exactly what the layer can and cannot prove: a
sound far-field single-step no-collision guarantee, per-give-way-step direction
compliance by construction, and a \emph{sound} imminent-unavoidability certificate
for the emergency regime. Upgrading these one-step guarantees, we prove that a
certificate-defined \emph{clearable set} is control-invariant, so a backup-maneuver
terminal constraint makes the shield \emph{forward-invariant} (recursively feasible)
on that set under a single constant-velocity target---a guarantee that reactive
CBF-QP filters lack under a $10$\,s zero-order hold---and on the shield's real
give-way state distribution a COLREGs-compliant certified escape exists at every
clearable state. We deliberately do \emph{not} claim bare provable collision freedom:
the guarantee is scoped to the clearable set (single CV), with the residual
($\approx\!0$--$3\%$, states clearable in no direction) deferred to an honestly
characterized emergency fallback. On the Krasowski--Althoff benchmark the shielded continuous policy attains
zero hull collisions over [PENDING: seeds$\times$episodes] and a structural
$\sim\!5\times$ advantage in throttle smoothness over the discrete baseline, at
comparable arrival. We report all limitations honestly, including a
training-stability collapse on a minority of seeds that we address with
policy warm-starting. [HONESTY: never "provably collision-free"; arrival is
comparable, not superior.]

% =====================================================================
\section{Introduction}
% [DRAFT skeleton]
Autonomous collision avoidance for surface vessels must satisfy the International
Regulations for Preventing Collisions at Sea (COLREGs), whose give-way rules are
\emph{directional} (e.g.\ in a head-on or crossing encounter the give-way vessel
must alter course to starboard). Two research lines have advanced maritime
collision avoidance but each concedes something essential:
\begin{itemize}
  \item \textbf{Discrete + masking} (Krasowski and Althoff 2024) obtains
  provable direction compliance by enumerating a discrete action set and masking
  non-compliant actions---at the cost of the continuous control that real actuators
  and smoothness require.
  \item \textbf{Continuous + soft reward} encodes COLREGs as a reward penalty,
  keeping continuous control but forfeiting any hard guarantee.
\end{itemize}
The intersection---continuous action $\cap$ provable direction compliance $\cap$
COLREGs---has been empty. We fill it with a projection-based safety shield adapted
from the safe-RL / action-projection literature to maritime COLREGs.

\paragraph{Contributions.}
\begin{enumerate}
  \item A projection-based COLREGs shield in the continuous action space
  (Section~\ref{sec:method}), integrated with RL in the safe-environment (SE-RL)
  style so the policy learns against the shielded dynamics.
  \item A \emph{provable safety layer} with an explicit, honestly-scoped set of
  guarantees (Section~\ref{sec:provable}): far-field single-step no-collision
  ($\forall$ target, sound), per-give-way-step direction compliance (by
  construction), and a sound imminent-unavoidability certificate for the emergency
  regime; extended to multiple targets ($\forall i$ / $\exists i$). We are explicit
  about what is \emph{not} proven.
  \item An empirical evaluation on the Krasowski--Althoff benchmark showing zero
  hull collisions and a structural smoothness advantage, with a candid account of a
  minority-seed training-stability collapse and its warm-start remedy
  (Section~\ref{sec:results}).
\end{enumerate}
% [HONESTY] We do not claim provable collision freedom or superior arrival rate.

% =====================================================================
\section{Related Work}
% [DRAFT skeleton · 待补引用]
\begin{itemize}
  \item \textbf{COLREGs-aware RL for USVs}: Krasowski and Althoff (2024) (discrete
  provably-compliant shield; our anchor and baseline); Meyer et al.\ (2020)
  (COLREG-aware RL).
  \item \textbf{Safe RL via action projection / shielding}: SE-RL and
  zonotope/QP action projection (Markgraf et al.\ 2026, serl-sprl); control
  barrier functions (CBF) and predictive safety filters as the guarantee mechanism
  we relate to.
  \item \textbf{Training-stability / exploration failure and remedies}: seed
  sensitivity in deep RL (Henderson et al.\ 2018; Agarwal et al.\ 2021, rliable);
  warm-starting / guided RL (JSRL 2022; AWAC 2020; RLPD 2023) and reverse
  curricula (Florensa et al.\ 2017), which motivate our collapse remedy.
  \item [PENDING: external baselines APF / VO-ORCA / CBF-QP for the Pareto
  comparison, Section~\ref{sec:results}.]
\end{itemize}

% =====================================================================
\section{Method}
\label{sec:method}
% [DRAFT skeleton]
\subsection{Problem setup and shielded MDP}
We adopt the Krasowski--Althoff CommonOcean setting: yaw-constrained point-mass
vessel dynamics (Eq.~\ref{eq:dyn} in Section~\ref{sec:provable}), continuous
control $u=[a,\omega]$, and the handcrafted two-vessel encounter benchmark. The
policy is trained in the SE-RL style, i.e.\ against the \emph{shielded} transition
so that it observes and adapts to the projection.

\subsection{The projection-based COLREGs shield}
At each step: (i) a rule state machine classifies the encounter regime
$\rho\in\{\text{far},\text{give-way(head-on/crossing/overtake)},\text{stand-on},
\text{emergency}\}$ and the compliant give-way half-plane; (ii) the compliant
half-plane and per-target collision-free constraints define the safe action set;
(iii) a quadratic program projects the policy command $u$ onto the nearest safe
action $u_{\mathrm{safe}}$; (iv) when the safe set is empty ($P=\varnothing$) in
the emergency regime, an empirical last-resort controller (Algorithm~1) minimizes
collision risk. [HONESTY: the emergency branch is empirical, not guaranteed.]

% ---- 可证明层正文（独立文件，含命题 1/2/3 + 多障碍 + 汇总表 + 失败机理段）----
\input{sec_provable_safety_layer}

% =====================================================================
\section{Experiments and Results}
\label{sec:results}
% [DRAFT skeleton · 大量 PENDING = 待烧卡]
\subsection{Setup}
Benchmark: Krasowski--Althoff CommonOcean handcrafted two-vessel scenarios
(head-on / crossing / overtake). Metrics (reported as IQM with bootstrap CI over
seeds, never cherry-picked): arrival rate, hull-collision rate, COLREGs violation
rate, and control smoothness (throttle-increment, yaw-increment, jerk). We report
two arrival criteria: strict (position + heading $\pm9.74^\circ$ + time, faithful
to the CommonOcean \texttt{is\_reached}) as primary, and position-only as a clearly
labeled looser complement. [HONESTY: dual criteria because Krasowski's published
maritime eval code is not in hand.]

\subsection{Main comparison (four arms)}
% [PENDING: headline four-way money chart on the balanced/overtaking dataset]
Baseline / soft-reward / discrete-safe / continuous-safe on the balanced dataset
(head-on + crossing + overtake). Expected honest positioning:
\begin{itemize}
  \item \textbf{Collision}: zero hull collisions for the shielded continuous arm
  (empirical; [PENDING seeds$\times$episodes]); safety methods tie near zero.
  \item \textbf{Throttle smoothness}: structural $\sim\!5\times$ advantage of the
  continuous arm over discrete (discrete has 7 quantized throttle levels).
  \textbf{[HONESTY] yaw-increment and jerk currently \emph{lose} to discrete
  ($\sim\!1.2$--$1.3\times$); de-jittering the yaw axis is open work.}
  \item \textbf{Arrival}: comparable to discrete, \emph{not} superior.
  [HONESTY: after balancing hidden collapsed seeds across arms, arrival shows no
  provable advantage over discrete (L192); we therefore do not list arrival as a
  selling point.]
  \item \textbf{Violation}: [PENDING same-budget same-dataset paired test.]
\end{itemize}

\subsection{Training-stability collapse and the warm-start remedy}
% [DRAFT · from 失败机理与紧急兜底刻画_0714 + L190/L191/L191-B/L192]
A minority of seeds (2/10 in the from-scratch golden run) collapse: the policy
never escapes a ``full-speed circling / time-out'' attractor and arrival flat-lines
near zero. We show (position-vs-heading decomposition; limit-cycle trajectories)
that this is a \emph{training-instability} limitation, not a criterion artifact,
and that reward shaping does not fix it (14 dosage sweeps). Warm-starting all seeds
from a single healthy policy escapes the bad basin (collapsed seeds' failure mode
changes entirely from time-out to near-goal stopping; zero hull collisions
maintained). [HONESTY: we describe this as ``warm-started training,'' never
``stable from scratch''; the ten warm-started seeds share one parent, so their
low spread is a fine-tuning band, \emph{not} a seed-variance claim---establishing
seed-level stability requires a multi-parent design, PENDING.]

\subsection{Emergency-regime characterization}
On an adversarial purely-geometric baseline that forces the emergency branch, we
observe the two collisions occurring with \texttt{source=emergency}, substantiating
that the emergency regime is an empirical fallback outside the far-field domain of
Proposition~\ref{prop:farfield}. [HONESTY: not a counterexample to the headline;
the headline is the provable direction-compliant shield with empirical zero
collision on the learned policy.]

% =====================================================================
\section{Discussion and Limitations}
% [DRAFT · 13 条 over-claim 收紧措辞已落此处 · 依据 03 L189 E + L191-B + L192]
% 数字来源：violation 1.13-1.8 / 崩率~20% CI[2.5,55.6] / 0-38 shared + 6.3-of-8 /
%   接管步 27-43% 路径~20% / 油门~5x 住 RL 段 / yaw 输 1.19-1.33x / 覆盖 96.32 vs 97.90.
\begin{itemize}
  \item \textbf{Provable tier is \emph{medium}.} gap\#1 (velocity-dimension
  unavoidability) is now sound, but coverage is not complete, and the empirical
  zero-collision backing is single-target until multi-vessel benchmark data exist.
  We keep every claim hedged and \emph{never} abbreviate the shield to ``provably
  collision-free.''

  \item \textbf{Compliance is per give-way step, not every step.} We claim provable
  \emph{direction} compliance only on give-way steps ($\sim\!13$--$16\%$ of
  encounters), never at every step (measured violation $1.13$--$1.8$ per episode on
  the golden run; the emergency regime structurally bypasses the constraint).

  \item \textbf{Arrival is comparable, not superior.} The apparent 24-point
  arrival win over discrete was a collapse-balancing artifact: after balancing the
  hidden collapsed seeds across arms, arrival shows \emph{no} statistically
  demonstrable advantage over discrete ($n=3$ per arm, sign-test $p=0.25$--$0.63$).
  We therefore do \emph{not} list arrival as a selling point.

  \item \textbf{The smoothness advantage is throttle-only and dilutes with
  takeover.} The $\sim\!5$--$6\times$ throttle-increment advantage is structural
  (discrete has 7 quantized throttle levels) but lives in the \emph{RL} segment; on
  the geometric-takeover segment the two arms are a wash, so the advantage shrinks
  as the takeover fraction grows. Yaw-increment and jerk currently \emph{lose} to
  discrete ($\sim\!1.19$--$1.33\times$); de-jittering the yaw axis is open work.

  \item \textbf{A minority-seed training-stability collapse.} A minority of seeds
  ($\sim\!20\%$, wide CI $[2.5,55.6]$) collapse. This is training instability, not
  scenario hardness: the shared-failure scenarios are $0/38$ unsolvable by all
  healthy seeds, and each is solved by $6.3/8$ healthy seeds on average (a clean
  seed-only control). Reward shaping is a \emph{high-confidence empirical} dead end
  for the tried shaping/coefficient families, \emph{not} a proof of impossibility
  for CMDP or non-PBRS terminal objectives.

  \item \textbf{Warm-start escapes the bad basin; it is not from-scratch
  stability.} All ten warm-started seeds derive from a single, post-hoc-selected
  parent, so their low spread is a fine-tuning band with effective sample size
  $\approx 1$, \emph{not} a seed-variance claim. Establishing seed-level stability
  requires the multi-parent design [PENDING]. Warm-start also slightly lowers
  provable-layer coverage ($96.32\%$ vs $97.90\%$), which we report and explain as
  a composition effect.

  \item \textbf{Geometric terminal takeover is a hybrid rescue, not RL learning.}
  Where used, early geometric takeover is a deployment-time rescue of the terminal
  docking failure (occupying $\sim\!27$--$43\%$ of steps / $\sim\!20\%$ of path for
  collapsed seeds); the RL policy itself is not improved. We describe it as a hybrid
  system, complementary to warm-start (which treats the full-speed-circling basin),
  not as ``the RL learned to dock.''

  \item \textbf{[PENDING]} multi-target empirical backing; a hard-dataset stress
  test to make the shield's value visible where the current benchmark is too easy;
  external CBF/VO Pareto baselines.
\end{itemize}

% =====================================================================
\section{Conclusion}
% [DRAFT]
We presented a projection-based, continuous-action COLREGs shield with an
honestly-scoped provable safety layer, filling the previously empty intersection of
continuous action, provable direction compliance, and COLREGs. We are explicit
about the boundary of the guarantee and about a training-stability limitation and
its warm-start remedy. Future work: multi-target empirical validation, a
multi-parent seed-variance study, yaw-axis smoothing, and external
(CBF/VO) Pareto baselines.

\section*{References}
% [PENDING: BibTeX; anchors: Krasowski \& Althoff 2024; Markgraf et al. 2026;
%  Meyer et al. 2020; Florensa et al. 2017; JSRL 2022; AWAC 2020; RLPD 2023;
%  Henderson et al. 2018; Agarwal et al. 2021.]

\end{document}
